\section{Compatibility with guilty pleas}

Looking at the procedure outlined in \textit{Criminal Code} s 606, it is clear that it explicitly and purposefully facilitates guilty pleas. Not only does \textit{Criminal Code} s 606 overtly permit defendants to plead guilty or not guilty, but it also explicitly excludes any other plea from being entered.\footnote{That is, apart from the ``special"/double-jeopardy pleas outlined in \textit{Criminal Code} s 607.} With the addition of \textit{Criminal Code} s 606(1.1), Parliament signalled that defendants should enter guilty pleas knowingly, voluntarily, and clearly understand the consequences of doing so. In order to help ensure that defendants do so, \textit{Criminal Code} s 606(1.1) outlines a series of subjective and objective requirements that must obtain for a judge to accept a plea:

\begin{itemize}
    \item \textbf{Subjective components.} The defendant is:
    \begin{itemize}
        \item Entering the plea voluntarily
        \item Aware that the plea formally admits the elements of the offence
        \item Aware of the nature and consequences of the plea
        \item Aware that plea bargains are not binding on the court
    \end{itemize}
    \item \textbf{Objective components.} The charge is:
    \begin{itemize}
        \item Supported by the facts
    \end{itemize}
\end{itemize}



The result is that a defendant who pleads guilty in Canada accepts responsibility for the offence charged. They agree that the allegations are true and acknowledge, usually implicitly, that the state has provided them with all of their procedural rights. They recognize that the judge who sentences them will ultimately determine a fit and appropriate sentence, notwithstanding any recommendations made by counsel. They also recognize that it is up to the court to determine whether it will accept their plea and that the judge must ensure that the plea was voluntary and informed. This process is what is commonly known as the ``plea inquiry." 



To the extent that we can define a guilty plea as an admission of the factual accuracy of the allegations and the procedural adequacy of the state's investigation of those allegations, we can say that the Canadian criminal framework is perfectly compatible with guilty pleas.