\thispagestyle{empty}

\begin{center}
    \subsubsection*{Acknowledgments}
\end{center}

\noindent
It would not have been possible to finish this thesis without the help and support of those around me, including family, friends, and colleagues. To my friends, family, and especially my husband Will, my sincere thanks for your time, interest, and patience throughout this process. In particular I would like to thank both my colleagues at Rees Dyck Rogala Law Office in Winnipeg, MB for their support and engagement in the early stages of the thesis, as well as my area director Adrian McBride at Legal Aid Saskatchewan in Swift Current, SK for the same in the later stages. Balancing work and academics is a collaborative effort, and I'm grateful for your help seeing this project through. In the same spirit, I truly appreciate the financial support from the Adalsteinn Fridrikson Kristjansson, David T. Sgayias, and Trevor Anderson Graduate Fellowships, which was critical in allowing me to balance work and scholastics.

I am also thankful for the insights and support that Dr. Donn Short provided during the coursework phase of my degree. Although we were relegated to remote learning throughout the entire coursework phase, Dr. Short nevertheless ensured that the LLM experience was a warm, welcoming, and supportive one. Similarly, professors Brandon Trask and David Ireland were both gracious enough to agree to review my work in its later stages and provide their valuable insights. Their perspectives were appreciated, provocative and astute, and helped focus and nuance my analysis.

Finally, I want to specially thank Dr. Amar Khoday for agreeing to supervise me throughout this endeavour. Dr. Khoday was a mentor and occasional employer for research projects throughout my JD program, and has been an ardent academic motivator and valued friend of mine ever since. He first suggested the LLM to me during a chance encounter at the airport one day, encouraged me to develop this research project thereafter, and has been supportive, enthusiastic, and generous throughout the process. I'm confident none of this would have happened without his help and collaboration, for which I am very grateful.