\section{\textit{Nolo contendere}}

The problems that plea bargaining critics have with the practice amplify with the \textit{nolo contendere} plea.\footnote{Critics like Bibas are a prime example. Bibas is not staunchly opposed to plea bargaining and believes the practice may have legitimate uses. But he is adamantly opposed to \textit{nolo contendere} and \textit{Alford} pleas, whose quirks and features he believes amplify the worst aspects of plea bargaining. See Stephanos Bibas, "Bringing Moral Values into a Flawed Plea-Bargaining System" (2003) 88:5 Cornell L Rev 1425 at 1432.} Even plea bargaining advocates are wont to disdain \textit{nolo contendere} pleas as ineffective, unreliable, and unjust.\footnote{See Bowers, \textit{supra} note XXX at 1165 - 1170.} The criticisms levelled against \textit{nolo contendere} pleas, in particular, may be tracked using the same problem criteria used to track plea bargains generally:

\begin{itemize}
    \item \textbf{The truth problem.} Defendants who self-convict while refusing to admit guilt or while actively maintaining their innocence say one thing while meaning another. Doing so is untruthful, and lawyers who assist clients with such pleas risk misleading the court by allowing defendants to deceive it openly. 
    \item \textbf{The fairness problem.} People charged with crimes are presumed innocent until proven guilty. The presumption should still stand when a defendant pleads guilty but protests their innocence or refuses to acknowledge their guilt openly. Where a defendant self-convicts but does not acknowledge their guilt, their motives are suspect. Such pleas thus appear unfair. 
    \item \textbf{The substantive problem.} \textit{Nolo contendere} pleas are evasive. Where guilty and not guilty pleas can signify both the truth or falsehood of the underlying allegations and a defendant's belief about the charges, \textit{nolo contendere} pleas cannot. Allowing defendants access to ambivalent pleas sends the message that actual guilt and innocence are secondary concerns, thus undermining criminal law's moral core. 
    
\end{itemize}

\subsection{The truth problem}

One impediment to allowing defendants to enter \textit{nolo contendere} pleas is cognitive dissonance. Cognitive dissonance presents when a person holds or is asked to hold two or more contradictory viewpoints simultaneously. A person who self-convicts but refuses to admit guilt appears to do just that. Convictions naturally follow where defendants plead guilty or are found guilty after trial. However, where defendants enter pleas whose proof-content is ``guilty," but whose truth-content signifies ``not guilty" or even ``innocent," those pleas raise a \textit{prima facie} problem of truthfulness. Although seemingly contradictory, I argue that this problem is superficial and that it fundamentally misunderstands formal pleas and admissions. When pleas and admissions are defined and understood more carefully and precisely, these apparent difficulties become unproblematic. I propose defining pleas and admissions as follows:

\begin{itemize}
    \item \textbf{Propositions.} Propositions are statements about the world that are capable of being true or false. In criminal legal proceedings, facts generally appear as allegations that may be proven, admitted, disproven, or remain unproven.
    \item \textbf{Facts, belief and proof.} Facts are statements about the world that are true. Facts and truth are evaluated objectively, such that facts must be objectively true. A person believes a proposition when they subjectively accept that it is true. A proposition is proven once a person is induced into believing it. Both are evaluated subjectively, in that beliefs and proof both ultimately rely on the person deciding a proposition's veracity.
    \item \textbf{Pleas and admissions.} In criminal proceedings, pleas are propositions that primarily convey general information about what, if anything, prosecutors will be required to prove at trial. Similarly, admissions are propositions that a court may accept as having been proven. Pleas and admissions may also convey objective truth and subject belief, though both are incidental to their proof-function.
\end{itemize}

\subsubsection{Propositions and facts}

I define \textit{propositions} as statements about the world that are capable of being objectively true or false. ``The sun rises in the east and sets in the west," ``the rain in Spain is rarely in the plain," and ``the earth orbits the moon" are all examples of propositions. In the criminal legal context, propositions commonly appear as \textit{allegations} that may be proven, admitted, or disproven. ``The victim consented to the fight," ``the defendant took reasonable steps to verify the complainant's age," and ``the prosecutor has demonstrated guilt beyond a reasonable doubt" are all propositions that are capable of being objectively true or false and are susceptible to proof and disproof.

Propositions are ubiquitous in language and as such may contain any number of components. The three that I have identified and focused on in this thesis are the truth, belief, and proof components, as these three are sufficient for understanding pleas and admissions and how they function in the criminal justice context. Propositions are true or false independent of whether anyone knows, understands, or acknowledges their truth or falsehood. Whether objective truths are subjectively acknowledged as such are matters of belief and proof.

\subsubsection{Facts, belief and proof}

\textit{Truth} and \textit{fact} are closely related terms that may be used interchangeably in specific contexts. For example, a proposition may be synonymously described as being ``the truth" or ``a fact."\textit{Truth} is a quality that all \textit{facts} possess, such that all facts are \textit{true},\footnote{The decision to primarily use the word ``truth" as an adjective and ``fact" as a noun is intentional, as it is more coherent to speak of multiple facts being true rather than multiple truths being factual.} and there are no untrue facts. \textit{A fact is a statement about the world that is objectively true}, regardless of whether one subjectively understands or believes it is true.\footnote{By ``world," I mean the entire realm of objects that have existed, do exist and will exist, as well as the relationships between those objects.} Statements like ``AlphaGo beat Lee Sedol in four out of five Go games at the DeepMind Challenge Match," ``the earth orbits the sun," and ``the fridge to my right is white" are all examples of facts in that they are all statements about the world that are objectively true. All facts are statements about the world, but not all statements about the world are facts. Statements like ``Lee Sedol beat AlphaGo in three out of five Go games at the DeepMind Challenge Match," ``the sun orbits Jupiter," and ``the fridge on my right is orange" are all false statements about the world, and therefore are not facts.

Whereas truth and facts relate to a proposition's objective veracity, \textit{beliefs} relate to a person's subjective acceptance of a proposition's veracity. A person believes a proposition when they accept that the proposition is true. A belief is true when it corresponds to objective reality and false when it does not.  

\textit{Proof} refers to grounding a subjective belief in ostensibly objective criteria. A proposition is proven when one subject satisfies another with justifiable reasons to believe it is true. Obversely, a proposition is disproven once the subject is satisfied that there are justifiable reasons to believe it is false or that there are no justifiable reasons to believe it is true. Where multiple subjects share similar or identical criteria, those criteria may form a common standard of proof. Proofs are generally founded on objective or external criteria but exist to satisfy the subject that their belief is reasonable.\footnote{See John Wigmore, \textit{The Principles of Judicial Proof} (Boston: Little, Brown, and Company, 1913) at 13 - 14: ``Proof ... is the result of that natural process of mind which all men would use in weighing the evidence that has already been admitted. Judges constantly find it necessary to warn us that their function in determining Admissibility is not that of final arbiters, but merely of preliminary testers, \textit{i.e.}, that the evidentiary fact offered does not need to have strong, full, superlative, probative value, does not need to involve demonstration or to produce persuasion by its sole and intrinsic force, but merely to be worth consideration by the jury. ... Proof, then, is obviously a distinct thing from Admissibility; because each evidential fact is offered separately, and we could not expect any one fact to produce demonstration. Since the production of evidential facts takes time, and one fact must precede another, we do not come to the question of Proof until all the evidence is in and the jury is ready to retire.}

True propositions may be proven false, while false propositions may be proven true. A proposition, once proven, may be disproven. For example, at various stages in history, it was proven that women had fewer teeth than men,\footnote{See Bertrand Russell's article.} that the earth revolved around the sun,\footnote{E.g., the Ptolemaic universe.} and that Thomas Sophonow murdered Barbara Stoppel. These propositions have been disproven, so none of these beliefs may be reasonably held today. Because these untrue propositions were once proven, \textit{there is no necessary correlation between proof and fact}. In order to properly understand how pleas and admissions function as proof-conveying devices, these concepts of truth, belief, and proof must be distinguished.

\subsubsection{Pleas and admissions}

A plea responds to a criminal allegation that a prosecutor must prove. Pleas are also propositions in that they contain a statement about the world that may or may not be accurate. However, pleas do not necessarily reflect a defendant's belief. A defendant may believe that they are guilty of an offence, may, in fact, \textit{be} guilty of an offence, but still plead not guilty. They may do so because the propositional content of a plea does not necessarily say anything about truth or belief. A defendant pleading not guilty does not necessarily provide the prosecutor or the court with any information about their beliefs or the truth of that proposition. Instead, they notify the prosecutor and the court that the offence's essential elements must be proven.

Similarly, a defendant who enters a \textit{nolo contendere} plea notifies the prosecutor and the court that the offence's essential elements need not be proven but does not necessarily provide any additional information about their subjective beliefs or the objective truth of those beliefs. Even where a defendant believes that the content of their plea is accurate, their subjective beliefs do not necessarily reflect the objective truth.\footnote{In the timeless words of Judge Maya Gamble to defendant Alex Jones, ``Your beliefs do not make something true."} However, in every case, a plea's proof-content is conveyed whenever a defendant enters one. Conveying this proof-content is a plea's most crucial function.

Admissions also advance propositions that do not need to be proven. In Canada, defendants may admit any ``facts alleged [by the prosecutor] for the purpose of dispensing with proof thereof."\footnote{See \textit{Criminal Code} s 655. If \textit{facts} signify only true propositions, the phrase ``facts alleged" is nonsensical, as \textit{allegations} may be true or false. The phrase is best understood as synonymous with a \textit{proposition}, and I will refer to it as such.} Typically, these allegations will support or consist entirely of an element of the offence charged. However, any mutually agreed upon proposition can be admitted under this section.\footnote{Defendants may wish to forego proof of some or all of the facts of their case for several reasons. In some cases, there is no reason to doubt that the prosecution will prove certain offence elements. A prosecutor is likely to be able to prove elements like date, time, jurisdiction, and identification in a domestic violence prosecution without difficulty. Agreeing to these facts saves court time and judicial attention, which can be minimal resources in busy court circuits. Admissions can also be advantageous in situations where the evidence is less certain to be proven. For example, when the prosecution wishes to call a witness whose evidence will inevitably harm the defendant to one degree, the defendant may wish to have the witness excused in exchange for a manageable set of agreed facts. In cases where the only genuine dispute lies with a legal issue, both parties may agree to have all evidence admitted by consent and limit trial time to arguing that issue.} Just as an uncontested plea obviates the need for the prosecutor to prove the offence's elements, admissions remove any need for either side to prove the proposition in question,\footnote{See \textit{R v Korski (CT)}, 2009 MBCA 37 at paras 119 - 128 for a discussion of the distinction between formal and informal admissions.} and once a defendant has admitted an allegation, otherwise admissible evidence may not be called or considered to contradict that admission.\footnote{See \textit{Handy}, \textit{supra} note XXX at para XXX: ``The issues in question derive from the facts alleged in the charge and the defences advanced or reasonably anticipated. It is therefore incumbent on the Crown to identify the live issue in the trial to which the evidence of disposition is said to relate. If the issue has ceased to be in dispute, as for example, when the fact is admitted by the accused, then the evidence is irrelevant and it must be excluded."} As with pleas, formal admissions are much less a matter of \textit{truth} and much more so a matter of \textit{proof}.

Pleas and admissions are primarily propositions about proof, not truth or belief. Consequently, the \textit{truth problem} is a definition problem. By entering a plea or making an admission, defendants advise the prosecutor and the court which propositions will require proof if any. Pleas are not evidence, and defendants neither swear nor affirm that the truth-contents of their pleas are genuine when they enter them. As a result, guilty defendants may plead not guilty, set a trial, and not worry about being liable for perjury. Where a defendant is unwilling or unable to enter a plea, the court must enter a not guilty plea on their behalf, regardless of whether the defendant \textit{believes} they are not guilty and whether the defendant \textit{is} not guilty. Where a defendant resiles their self-conviction, Canadian courts focus on whether the defendant entered the plea voluntarily, knowingly, and unequivocally, and not on whether the plea was factually true.\footnote{See e.g. \textit{R v Behr}, 1966 CanLII 252 (ON SC). In that case, the Ontario Superior Court upheld a guilty plea despite the defendant being told by his lawyer that the guilty plea was not an admission of guilt, as it was nonetheless satisfied that the plea was voluntary. But there are exceptions to this. See \textit{R v Catcheway}, 2018 MBCA 54, where the defendant pleaded guilty to being unlawfully in a dwelling house despite having been incarcerated at the time. The Manitoba Court of Appeal allowed him to withdraw his guilty plea because there were ``valid grounds for doing so." Absent evidence that his pleas were uninformed or involuntary, it erred in doing so. To its credit, the court simultaneously arrived at the correct solution by allowing the appeal against conviction based on a miscarriage of justice, as there was compelling evidence that the defendant was innocent.} A plea's primary purpose is to signal to the court whether evidence will be called or contested and not to convey a defendant's actual beliefs or the objective truth about those beliefs.

\subsection{The fairness problem}

Although \textit{nolo contendere} pleas are not implicitly untruthful, there is still an intuitive sense that they are unfair. Defendants may enter a guilty plea because they sincerely believe the allegations or wish to take responsibility and suffer a deserved consequence. Pleas that a defendant enters in these or similar circumstances seem fair on their face. By contrast, a defendant who enters a \textit{nolo contendere} plea does not take responsibility for the offence. This fact makes it unlikely that a defendant will enter the plea because they believe it is accurate or deserves the consequences. Pleas that a defendant enters in these or similar circumstances seem unfair on their face.

As discussed above, Canadian criminal law has a lengthy history of allowing defendants to plead guilty only when they do so knowingly, voluntarily, and unequivocally. These principles are so deeply ingrained in Canadian criminal law that they found their way into the informal \textit{nolo contendere} plea procedure and are a fundamental feature of Canadian guilty pleas.\footnote{Similarly, in most American jurisdictions that allow \textit{nolo contendere} pleas, the plea voluntariness and comprehension inquiry required for inculpatory guilty pleas is often mirrored for non-culpatory \textit{nolo contendere} pleas.} As a result, any time a defendant invites a conviction, the court receiving their plea must ensure that they are not doing so due to undue pressure, short-sightedness, or confusion. But despite these protections, some concerns may remain. Where a defendant is presumed innocent and does not admit that they are guilty, it is reasonable to suspect that something unfair has transpired. Answering this argument requires looking at both the \textit{factual foundation} requirement and the residual \textit{judicial discretion} to refuse to accept a self-conviction.

\begin{itemize}
    \item \textbf{Factual foundation.} Canadian judges may only allow a defendant to plead guilty if there is a factual foundation for the charges. This judicial oversight helps ensure that defendants do not convict themselves when faced with unfounded or unlawful accusations. Insisting on this requirement for \textit{nolo contendere} pleas helps ensure that defendants are not held liable for unfair or untruthful allegations.
    \item \textbf{Judicial discretion.} Defendants cannot self-convict as a matter of right. As a result, even where Canadian judges are satisfied that allegations are factually founded, they may still reject a defendant's guilty plea. Giving judges final discretion over whether to accept a plea allows them to fulfill their judicial mandate to treat defendants fairly.
\end{itemize}

\subsubsection{Factual foundation}

Notwithstanding that defendants can legally admit facts against their interests and invite the court to convict them, there are situations where it would be unjust for the court to do so. One such situation is where the propositions the prosecution relies upon do not make out the offence charged. Where a defendant formally admits the offence but otherwise maintains their innocence, questions reasonably arise as to whether that condition obtains. To help ensure that defendants do not invite baseless convictions, the \textit{Criminal Code} requires that judges ensure that ``the facts support the charge."\footnote{See \textit{Criminal Code}, \textit{supra} note XXX s 606(1.1)(c).} Although the common law does not require judges to ensure that there is a factual foundation for \textit{nolo contendere} pleas,\footnote{See Pickle, \textit{supra} note XXX at 250 - 251.} many American jurisdictions have condified the requirement into law.\footnote{States that have legislated this requirement include Arizona, Arkansas, California, Colorado, Delaware, Florida, Kansas, Maine, Massachusetts, Michigan, New Hampshire, North Carolina, Ohio, Oklahoma, Rhode Island, Utah, and Wisconsin. Similarly, best-interest pleas like \textit{Alford} may only be entered when judges are satisfied that a factual foundation exists.}

Where a defendant either refuses to accept responsibility for an offence overtly or actively protests their innocence, a review of the allegations and the supporting evidence lets the court evaluate their apparent equivocations. In cases where the allegations are straightforward, the evidence supporting the allegations is strong, or both, defendants who vacillate after self-convicting may be safely ignored. In cases where the allegations are complex, or the evidence supporting the allegations is equivocal, judges may consider these factors when deciding whether to allow the defendant to convict themselves or exercise their discretion and set a trial date.

Requiring a factual foundation for both guilty and \textit{nolo contendere} pleas reflects an appropriate concern for fairness. Defendants who enter explicit guilty pleas may reassure the court to some degree that they understand and admit the allegations they face. Where defendants cannot or will not enter explicit guilty pleas but nonetheless wish to self-convict, the court may nonetheless assure itself that the allegations are accurate and the charges stemming from those allegations are fair by reviewing them before accepting a plea. But even where the allegations support the charge, this judicial oversight may put a stop to pleas that are otherwise unfair.

\subsubsection{Judicial discretion}

In Canada and the United States, judges have immense discretion when deciding whether to accept or reject most uncontested pleas. Among jurisdictions that formally allow \textit{nolo contendere} plea, only one state allows defendants to enter that plea by right. This broad discretion is a final safeguard for a defendant's presumption of innocence. Even where a court is satisfied that a defendant is inviting a conviction and that there is sufficient evidence to support convicting them, a judge may still refuse to authorize the plea. Because criminal defendants generally cannot self-convict by right,\footnote{See e.g. \textit{R v Goodheart}, 2005 ABCA 126 at para 2; \textit{R v Turnbull}, 2016 NLCA 25 at para 4; R v Podolski, 2018 SKPC 13 at para 1.} the presiding judge is under no obligation to give reasons for doing so, but this role and power should be directed at preventing pleas that the judge suspects should not result in convictions. Neither defendants nor prosecutors have eminent domain over the presumption of innocence. Rather, defendants are presumed innocent until a judge is satisfied that they are guilty.

Defendants are presumed innocent throughout the criminal process but may waive that presumption and invite a conviction. However, the presumption still stands until a court is satisfied that there is a reasonable basis for allowing that conviction. Like any other person, criminal defendants are susceptible to ``buyer's remorse," minimizing or denying their behaviour, and misunderstanding polysemic legal concepts.\footnote{The problem of correctly understanding abstract legal concepts may be insoluble. See e.g. Frederick Schauer's ``Formalism" at 514: ``Some terms, like `liberty' and `equality' are \textit{pervaseively indeterminate}. It is not that such terms have no content whatsoever; it is that \textit{every} application, every concretization, every instantiation requires the addition of supplementary premises to apply the general term to specific cases."} Where a defendant has told the court that the prosecution does not need to prove its case against them, and where there is an adequate allegational matrix to support the prosecution's case, the presumption of innocence no longer stands. It is difficult to convincingly argue that proceedings are unfair when defendants are both presumed innocent and where judges must ensure that defendants only self-convict on demonstrable facts that support the charge.

\subsection{The substantive problem}

Bibas broadly argues that the contemporary common law criminal justice system has a hyper-inflated focus on procedure that erodes its ability to achieve its underlying moral aims.\footnote{See Bibas, \textit{supra} note XXX at 1362: ``Criminal procedure has for too long treated itself as a subset of constitutional law, 1 in the process distancing itself from substantive criminal law. Although substantive criminal law sometimes discusses how well rules deter, rehabilitate, or exact retribution, these substantive values are largely absent from criminal procedure. Instead, criminal procedure seems to care only about whether procedures are efficient, constitutional, fair, and accurate."} Although his argument applies to expediently-entered and insincere guilty pleas by extension, Bibas's criticisms are primarily focused on the ``ambivalent" \textit{nolo contendere} and \textit{Alford} pleas.\footnote{See Bibas, \textit{supra} note XXX at 1403.} Bibas's \textit{nolo contendere} and \textit{Alford} plea criticism tracks his broader rebuke of efficiency-driven plea bargaining. He argues these pleas should be scrapped as they encourage defendants to avoid confessions, undermine the law's moral authority by allowing defendants to cling to their own dissonant values, and bypass the curative trial process to universal detriment. I argue that Bibas's concern for substantive criminal law is compelling but ultimately misdirected, and answer his criticisms under three general headings:

\begin{itemize}
    \item \textbf{Accepting valuable compromises instead of confessions.} Sincere confessions may be preferable to ambiguous admissions, but that does not mean the latter have no value. \textit{Nolo contendere} pleas help secure otherwise unobtainable convictions, mete out punishments to defendants willing to accept them, and allow more defendants to self-convict with a clear conscience.
    \item \textbf{Exposing ideological divides and explicating educational opportunities.} When defendants are allowed to accept the benefits of plea bargaining while refusing to admit guilt outright they expose an ideological divide between themselves and the state convicting them. Once this divide is explicit, courts have a uniquie opportunity to investigate its causes and educate those whose private morality is at odds with public standards.
    \item \textbf{Avoiding the harms unwanted trials cause.} Defendants who cannot or will not plead guilty must set a trial. This exposes defendants and all justice system participants to the potentially harmful effects that trials entail and opens the proceedings up to wrongful acquittals.
\end{itemize}

\subsubsection{Avoiding confessions through ambivalent pleas}

As discussed in \S 4.1.3 above, Bibas places great value on confessions. I argued there that Bibas overvalues confessions by pointing to the clear value of obtaining otherwise unobtainable convictions, the unclear value of obtaining confessions without a way to ensure that they are sincere, and the lack of data to quantify the rehabilitative value of even sincere confessions. Rehabilitation will always reasonably favour a receptive offender. But without any logical reason or compelling evidence demonstrating that restricting \textit{nolo contendere} or other ambivalent pleas correlates to an increase in sincere guilty pleas, they should not be limited.

But even if Bibas overvalues confessions, this does not mean that his concerns about ambivalent \textit{nolo contendere} pleas are misplaced. Sincere confessions are valuable, even if they are not as valuable as Bibas believes them, and cynically-entered \textit{nolo contendere} pleas can undermine substantive values. While plea bargains may convince some guilty defendants to admit culpability, pleas like \textit{nolo contendere} arguably allow defendants to avoid taking responsibility for their offending behaviour. 

As a result, critics like Bibas may remain open to plea bargains while condemning non-inculpatory self-convictions. This position allows Bibas to occupy a middle ground where  plea bargaining is encouraged, as long as defendants enter unambiguous pleas. Bibas argues that the harm caused by plea negotiations come from defendants being allowed to avoid taking responsibility. \textit{Nolo contendere} pleas allow defendants to avoid confessions, thereby avoiding the rehabilitation that confronting difficult truths about themselves and their actions would achieve. Bibas specifically highlights cases where defendants are ordinarily reluctant to confess\footnote{Sexual offences generally, sexual offences against relatives or children specifically, and other socially heinous crimes fall into this offence class. See Bibas, \textit{supra} note XXX at 1393 - 1395.} as prime examples of cases where courts, prosecutors, and defence lawyers \textit{should} compel them to plead guilty rather than encouraging plea deals or permitting them to plead \textit{nolo contendere}.\footnote{Bibas does not mention the rehabilitative prospects of defendants convicted after trial.} This sharply contrasts with the conventional wisdom that encourages these pleas as reasonable compromises to resolve difficult cases.

Bibas recognizes that punishment can play a part in reforming and deterring by educating them and society.\footnote{See Bibas, \textit{supra} note XXX at 1390.} But the fact that \textit{nolo contendere} pleas specifically invite punishment, and may thus play a part in reforming and deterring offenders under this definition, is at odds with Bibas's view that \textit{nolo contendere} pleas undermine society's moral authority. Moreover, defendants who enter these pleas put the court on notice that there is a divide between their moral norms and society's. To the extent that the court ought to pursue substantive criminal values on sentencing, they are better positioned to deal effectively with a defendant if they know this divide exists. Defendants who enter insincere guilty pleas or plead not guilty send inconsistent signals to the extent that they send any signals. \textit{Nolo contendere} pleas, on the other hand, notify the court that the defendant is willing to accept punishment but unwilling to admit that they are guilty, even though doing so would potentially mitigate their sentence. This should prompt the substantively-inclined court to ask why, and allows the court to investigate and try to bridge that divide.

Enabling more defendants to self-convict and safeguarding the means for them to do so provides defendants with more opportunities to be punished, thereby increasing the number of defendants who can be reformed, deterred, and educated. It is reasonable to infer that the results will be less optimal than those obtained through sincere guilty pleas. But it is also reasonable to infer that the results will be no less optimal than those obtained through insincere guilty pleas, and much more optimal than if the defendant is acquitted at a trial. Absent a reliable way to ensure sincere guilty pleas, defendants willing to self-convict should be allowed to do so regardless of whether they feel they deserve to be punished for their actions.

% Tie into ideological divides

More importantly, these pleas may be truthful in ways that a guilty or not guilty plea would not be. One of the law's intractable difficulties is that it relies on equivocal concepts and assumes that judges and juries use these concepts reasonably. In criminal law, \textit{excuse} and \textit{justification} are two particularly difficult examples. For example, what Parliament and the courts define as a reasonable amount of force in an assault will invariably differ from how individual defendants view their own actions. Justifications that individual defendants invoke to explain their actions may or may not coincide with legally valid excuses and defences. Once confronted by this reality, a defendant who accepts a conviction without expressing remorse or even agreeing with all of the underlying allegations accepts the law's right to punish them for their actions and the truth of their own personal convictions. This allows defendants to enter truthful pleas and exposes the ideological divide between these defendants and the state's broadly held values.

\subsubsection{Ideological divides}

Defendants willing to self-convict may hesitate to enter a guilty plea for many reasons. One reason stems from the moral divide between individual defendants who commit offences and the state that prosecutes them. Bibas identifies this phenomenon but hastily treats it as a fundamental problem. I argue that these pleas can helpfully identify defendants whose morality is out of sync with society's broader values, thereby providing prosecutors, judges, community corrections officers, and other actors in the justice system with information that may assist with their rehabilitation.

Where defendants enter ambivalent pleas like \textit{nolo contendere}, Bibas contends that ``society loses the authoritative vindication of its norms and the repudiation of the wrong."\footnote{See Bibas, \textit{supra} note XXX at 1407.} While it is problematic when the state and its citizens have competing or oppositional values, ambivalent uncontested pleas do not \textit{create} those differences, but merely \textit{expose} them. Rather than undermine the state's moral authority, as Bibas suggests, these pleas give defendants a forum where they can acknowledge the state's authority, notwithstanding their private differences. These pleas also allow the state to gauge a defendant's rehabilitation prospects using more accurate, albeit less encouraging information. Forcing a defendant to plead guilty as their only route to self-conviction may offer some empty assurances that the defendant will come to accept their guilty pleas as true,\footnote{See Bibas, \textit{supra} note XXX at note 15; see also \textit{ibid} at 1399: ``Some defendants are willing to confess and plead guilty. As Alschuler rightly notes, most guilty pleas are not the fruit of genuine repentance. EH Instead, defendants feign repentance to earn sentence reductions. But even feigned or induced repentance may teach lessons to some offenders. The very act of confessing and pleading guilty in open court heightens the defendant's awareness of the victim's injury, the norm violate d, and the community's condemnation. 1 \H-i Indeed, the ordeal of feigning repentance, even if initially done for the wrong reasons, can sometimes lead to genuine repentance."} but in fact ensures that the court will receive no useful or accurate information about what the defendant believes to be true. Instead, the interests of justice are better advanced by showing the public that justice prevails over private opinion and morality. A defendant's prospects for rehabilitation can be better understood when these belief differences are explicated, thus advancing the substantive values of reconciliation, mutual responsibility, and accuracy.

Allowing defendants who neither feel guilty nor remorseful to accept their legal culpability allows them to use uncontested pleas honestly, convey their beliefs effectively, and identify their rehabilitative prospects to the court more accurately. Once the court and others tasked with rehabilitating offenders understand what a defendant believes to be true they can better understand the circumstances of the offence and offender. Accurately identifying defendants who are not ideal rehabilitation candidates can help judges tailor fit and appropriate sentences, help probation officers identify clients who will require extra or special focus and alert law enforcement officers to potential community risks. Forcing such defendants to enter insincere guilty pleas or run trials negates these benefits and invites the other ill effects accompanying these unwanted contested hearings.

\subsubsection{The harms unwanted trials cause}

In \S 4.1.3 above, I examined Bibas's portrayal of trials as morality plays, and argued that this conceptualization is at odds with the adversarial justice system and inevitably leads to unjust results. Modifying the criminal justice system in the manner he suggests would curtail the defence lawyer's role as advocate, discourage resolution discussions between counsel, and inevitably result in wrongful convictions. In this section, I argue that Bibas's concept of trials as morality plays uncritically aggrandizes contested hearings as unqualified goods. In doing so, I argue that Bibas ignores the harms that trials can cause generally, and the harms that unwanted trials can cause specifically. To the extent that expanding permissible self-convictions to include \textit{nolo contendere} allows more defendants, victims, and other justice system participants to avoid these harms, they should be encouraged.

Bibas correctly points out that pleas like \textit{Alford} and \textit{nolo contendere} are likely to be less satisfying for victims and less conducive for rehabilitating offenders than a truly remorseful guilty plea. Unless defendants willingly admit guilt, Bibas proposes they should be forced to set a trial. He views this outcome as a win-win, in that defendants either must confront their wrongdoing, thus availing themselves of the curative properties of confession and punishment, or be taught a lesson through the evidence presented at their morality play of a trial.\footnote{See Bibas, \textit{supra} note XXX at 1403 - 1404.} But Bibas ignores the very real possibility that some defendants who are forced to set trials instead of self-convicting will be acquitted. Trials can be complex affairs and criminal charges must be proven beyond a reasonable doubt. Problems with securing witnesses or ensuring that all the necessary evidence is admissible are not uncommon, and even strong prosecutorial cases can unexpectedly collapse, often resulting in last-minute plea dealings and fire-sale sentencing discounts. 

As Bibas fails to consider the costs of wrongful acquittals and frantic eleventh-hour dealing, he fails to properly evaluate the benefits of sincere confessions and satisfying convictions. It is reasonable to assume that victims will be less satisfied with a non-commital \textit{nolo contendere} plea than a sincere guilty plea, but it is equally reasonable to assume that they will be less satisfied with a wrongful acquittal than they would have been with a non-committal conviction. Similarly, a lackadaisical \textit{nolo contendere} plea to a substantive offence with a victim is likely to be more satisfying for victims of that crime than a sincere guilty plea to an administration of justice offence a defendant picked up while awaiting trial. 

Wrongfully acquitted defendants who would otherwise accept a conviction may not generate the same outrage that wrongful convictions do, and rightfully so. However, they still undermine confidence in the justice system and subvert its substantive goals. Although much less harmful than wrongful convictions, wrongful acquittals are untruthful results that leave the guilty unpunished. Meanwhile, factually guilty defendants willing to resolve are required to set their matters down for trial and may spend more time on release conditions or confined to pre-trial detention than if they had been able to enter \textit{nolo contendere} pleas. It is dubious whether the defendant will be more prepared to sincerely admit guilt after hearing the matter tried than when they would have agreed to self-convict before accumulating these process costs. And likely, the unnecessary expense of the trial and the burdens that the hearing imposed on the witnesses will be on the sentencing judge's mind.

Requiring defendants to set trials when none of the parties involved want one ensures that there is one less trial date available for a truly contested matter. Society does not have a legitimate interest in requiring unwilling and otherwise cooperative defendants, victims, and professional justice system participants to engage in these literal show trials. Defendants who must have their allegations adjudicated, even where they lack viable defences, are forced to spend unnecessary time in pre-trial detention or constrained by pre-trial release conditions. At the same time, defendants with a viable defence are similarly delayed and imposed on. Those defendants willing to accept a final judgment to end these restrictions should be permitted to do so. Victims should be allowed to see punishment meted out, restitution made, and the process concluded, even if that sentencing package does not include a repentant defendant. 