\section{The cognitive dissonance of non-inculpatory no-contest pleas}

One impediment to allowing defendants to enter non-inculpatory no-contest pleas is the cognitive dissonance experienced when trying to hold the conflicting values required to sustain them. Having a person plead guilty only to protest their innocence is superficially contradictory, in that they say one thing (or nothing at all) while accepting the consequences of its opposite. In this sense, no-contest pleas that either deny guilt outright or refuse to admit it appear equivocal. This apparent equivocation is problematic for several reasons:

\begin{itemize}
    \item \textbf{The truth problem.} Canadian criminal law requires defendants who cannot or will not plead guilty to set their cases for trial. People consider trials a search for the truth where all evidence may be looked at and examined. By contrast, they see negotiated plea deals resulting in equivocal pleas as an obfuscation of the truth.
    \item \textbf{The presumption of innocence problem.} People charged with crimes are presumed innocent until proven guilty. Where a defendant pleads guilty but protests their innocence, they should be presumed innocent and not be convicted. Similarly, where a defendant refuses to admit to the charges, their innocence should still be presumed, and they should not be convicted.
    \item \textbf{The fairness problem.} Where a defendant invites a self-conviction and either does not admit guilt or actively protests their innocence, it is reasonable to suspect their motives. Because these defendants disclaim their guilt one way or the other, the pleas are not overtly products of good motives, such as a desire to take responsibility for their actions or to ensure that the truth wins out. Such pleas thus appear to be products of unfairness. 
\end{itemize}

\subsection{The problem of truth}

Convictions naturally follow where defendants plead guilty or are found guilty after trial. However, where defendants cannot or will not do so, but nonetheless insist that the court treat them as though they did, they appear to contradict themselves. As the content of the plea is effectively ``guilty," but the content of the defendants actions otherwise is ``not guilty" or even ``innocent," these pleas raise an apparent problem of truthfulness. As this section will demonstrate, this problem is merely superficial, and arises from a misunderstanding of the relationship between facts, proof, and truth on one hand, and the relationship of each to formal pleas and admissions on the other.

\subsubsection{Facts, anticipated facts, and proven facts}

To navigate the problem of truth, it is necessary to define terms. If non-inculpatory no-contest pleas raise a problem of truth, it is essential to define what ``truth" means. 

\textbf{Facts.} A fact is a statement about the world that is objectively true. 

\textbf{Anticipated facts.} In legal proceedings, parties to an action will propose ``statements of facts" to the court in one form or another. These may be presented orally or writing, and will very often contradict one another partly or entirely.\footnote{Parties to a legal proceeding do not necessarily have to disagree about any of the underlying facts. Trials can be run solely on the basis of a dispute about a particular law's application. See, for example, \hl{that case I discuss above}.} For example, in most criminal prosecutions that are set for trial, the defendant will advance and rely on a set of facts that 

\textbf{Proven facts.} A proven fact is a statement that a judge or jury accepts as having met the burden of proof. 

Throughout this thesis, ``facts" are understood as states that may or may not obtain in the world. 

\subsubsection{Correlation between facts, admissions, and pleas}

\begin{itemize}
    \item \textbf{Admissions.} Copy some of that goodness from below.
    \item \textbf{Pleas.} By default, a not guilty plea requires the prosecutor to prove all of the facts against a defendant. By default, a no-contest plea relieves the prosecutor of this burden.
    \item \textbf{Facts and admissions.} 
    \item \textbf{Facts and pleas.}
\end{itemize}

\subsubsection{The apocryphal truth-seeking function of the trial}

Because defendants who are unwilling or unable to offer a guilty plea must set their matters down for trial, a trial may be considered the ``default mode" for resolving criminal disputes. Because trials take place when parties disagree with one another, they are often branded as a ``search for the truth." Each party has the opportunity to present its best evidence and its best interpretation of the evidence. Witnesses are cross-examined on their evidence, identifying the strengths and weaknesses of their testimony, and a neutral trier hears the facts, listens to both sides argue, and decides between them.

By contrast, non-inculpatory no-contest pleas appear to be designed to obfuscate or misrepresent the truth for the purpose of getting negotiated agreements through the courts. Negotiations between prosecutors and defence lawyers are privileged and inadmissible as evidence. There are no judges or juries to provide neutral adjudication.\footnote{This is not necessarily true in all cases. In many jurisdictions, contested matters may go through case management with a judge in order to discuss pre-trial issues, canvass resolution, and offer advice on how to best proceed. }

Defendants can privately (or publicly) resile their admissions without being contradicted by evidence heard at a public trial

But the function of a common-law criminal trial is not to "search for the truth." Rather, the function of a common-law criminal trial is to allow the trier of fact to make decisions based on curated sets of facts, within the confines of a restrictive sets of rules. There are important differences between these two concepts. 

Seeing a trial as a "search for the truth" implies that truth is the primary function of the trial, which subsequently implies that truth is the value in a trial that all other values are subservient to. A public inquest into a social problem, like the Truth and Reconciliation hearings in ZA, or missing and murdered indigenous women in Canada, would be comparable examples of this sort of "no holds barred" approach to the truth. 

Trials, on the other hand, require the fact finder to consider a less complete and arguably less true version of the evidence, as in cases where evidence has been excluded. In criminal trials, the "truth-seeking function" of the trial is always counterbalanced by the need to ensure fair proceedings. Truth is, in fact, severely curtailed at trial by things like what truth the trier of fact is allowed to hear, excluded evidence, excluded testimony, the results of pre-trial and other evidentiary motions, testimony from witnesses who were not called or did not attend, hearsay rules, opinion evidence rules, and so on. Even once the evidence is admitted, special instructions are often needed to ensure that the evidence is interpreted properly. 

Practically speaking, a great deal of truth that could be presented at trial is often excluded simply for the sake of a coherent presentation

In all cases, but especially complex ones, lawyers looking to present their evidence to a judge or jury will regularly cut out large swaths of the available evidence

A lawyer's job at a common law criminal trial is to advocate for a position

Effectively advocating for a position requires the advocate to account for the frailties of human attention and comprehension

This in turn requires both extensive editing of the truth and careful framing of that truth within the context of a larger argument

Information at trial is highly pre-processed before it reaches the court, and needs to be further processed once it gets there. This ensures both procedural fairness and effective presentation of the truth.

Procedural fairness and presentation are key trial concerns because the truth-seeking mechanism is unreliable

The "finders of fact" at trial are no more implicitly qualified at "seeking truth" than anyone else

In the case of jurors, the "finders of fact" are, in effect, "anyone else"

Experience

The hope and expectation is that the fact finders will come to true and fair conclusions through the combined presentation of true and fair evidence

But this isn't guaranteed

Rules have been put in place following decades of learning about the frailties of evidence

DNA evidence created a new class of cases for courts to review

The theory is that certain convictions could be absolutely confirmed or disconfirmed based on how DNA tests came back

Where wrongful convictions were found, we were able to gain some insights into the types of evidence that led jurors to those conclusions

Other forensic experiments in other fields have confirmed the unreliability of certain types of evidence

Eyewitness evidence

Shortcomings in memory

Etc

More confirmed wrongful convictions appear to have resulted from trials rather than negotiated sentences

But until we have new breakthroughs that can offer insight into how well or poorly forensic evidence reflects the truth, there isn't another independent truth-seeking function at play

Other truth-seeking mechanisms, like resolution discussions and plea bargaining, are likely better suited for reaching outcomes that are both true and fair

Both (sets of) lawyers in a criminal action have intimate familiarity with the facts and nuances of the case. They represent the different interests in the action - or at least, the interests that have standing in the action. Assuming ethical actors with access to relatively complete and accurate information, any resolution worked out between the parties by consent is likely to be close to a true and just result.

This also assumes there's some *quid pro quo* between the parties in their resolution discussions, such that the evidence is already being "tested", in a sense

Where a resolution can't otherwise be reached, trials are a reasonably effective means of adjudicating disputes, ensuring all sides have their best explanation of the evidence heard, and deferring to an ostensibly neutral third party to decide between competing narratives.

But trials are probably better understood as adequate alternatives to dispute resolution, rather than a preferred means of dispute resolution

Negotiations between counsel, both of whom have legal training, are duty-bound to advocate for their positions, have access to the fullest permissible truth of a criminal matter, and the opportunity to spend extensive time and resources investigating the nuances thereof, are a much greater truth-seeking device than a criminal trial.

\subsubsection{Truth and validity do not necessarily correlate to formal admissions}

A formal admission does not need to be factually accurate or ``true" in order to be valid. 

Similarly, though arguably to a lesser extent, there is a dissonance when a person refuses to admit that they are guilty, only to be found guilty without a trial or evidence.

There is, therefore, the potential for a cognitive dissonance that will need to be addressed for some before equivocal no-contest pleas become a viable topic for discussion.

Trials are not an unqualified search for the truth and may not be the best means of obtaining the factual truth in criminal proceedings.

\subsubsection{Equivocal no-contest pleas}

Similarly, on the surface, equivocal no-contest pleas appear to be duplicitous.

The defendant either pleads guilty or, at the very least, refrains from declaring that they are not guilty.

However, despite either declaring guilt or refusing to declare that they are not guilty and being convicted of the offence, the defendant may later state that they are factually innocent of the charge.

On the surface, these no-contest pleas appear to be duplicitous.

A person appears to be claiming one thing while claiming something else.

Alternatively, they withhold information that, if true, ought to be admitted into evidence.

This apparent conundrum may be reframed in terms of the need to ensure that the court is not misled, and counsel's role in that respect

Lawyers are under a strict duty to not intentionally mislead the court

This includes calling evidence at a trial that the lawyer knows to be misleading

A common example is the lawyer who argues that a third party committed the crime after the defendant privately admitted their guilt

See the SK Code of Professional Conduct at Section 5.1-1 (10)

Admissions made by the accused to a lawyer may impose strict limitations on the conduct of the defence, and the accused should be made aware of this. For example, if the accused clearly admits to the lawyer the factual and mental elements necessary to constitute the offence, the lawyer, if convinced that the admissions are true and voluntary, may properly take objection to the jurisdiction of the court, the form of the indictment or the admissibility or sufficiency of the evidence, but must not suggest that some other person committed the offence or call any evidence that, by reason of the admissions, the lawyer believes to be false. Nor may the lawyer set up an affirmative case inconsistent with such admissions, for example, by calling evidence in support of an alibi intended to show that the accused could not have done or, in fact, has not done the act. Such admissions will also impose a limit on the extent to which the lawyer may attack the evidence for the prosecution. The lawyer is entitled to test the evidence given by each individual witness for the prosecution and argue that the evidence taken as a whole is insufficient to amount to proof that the accused is guilty of the offence charged, but the lawyer should go no further than that.

Though superficially convincing, and seemingly often used, this analogy falls apart in a few ways

It either overestimates, or fails to account for how rarely a lawyer actually has the opportunity to **knowingly** or **intentionally** mislead the court in criminal matters

Many lawyers take steps to avoid learning this information in the first place

Those who don't will likely agree that they rarely have enough knowledge, or enough confidence in the knowledge imparted to them by their client, to knowingly or intentionally tell the court much of anything about the case

Like Stephanos Bibas, it recasts the defence lawyer's role as something other than it currently is

Stephanos Bibas also wanted to see defence lawyers take up a more inquisitional role, but for different reasons

Some investigation and thought is obviously required on a defence lawyer's part before they can enter pleas for their client, especially as the client may not realize they have a defence

But for a client whose clear and unambiguous instructions are to enter a guilty plea, there's only so much investigation and interrogation that a defence lawyer should be doing

There's also some question about whether one can convincingly say that a lawyer who hasn't done their due diligence in this manner is intentionally misleading the court

Unless the lawyer actively adduces facts that aren't true on behalf of their client, assisting them with entering a plea isn't misleading the court

The simple plea of "guilty" isn't sufficient, as will be explored immediately below

But that apparent contradiction is only superficial, the fact of which is borne out when examining the absence of any apparent contradiction with not guilty pleas

\subsubsection{Counterexamples}

\paragraph{The not guilty plea\\}

The same apparent contradiction does not appear to hold true when considering a defendant who declares that they are innocent, pleads such, is subsequently convicted after trial, but continues to declare that they're innocent

Part of the lack of apparent contradiction comes from the fact that a trial has been held

The defendant has been given their chance to prove that they weren't guilty, as it were, and wasn't able to do so

One problem with this approach is that it ignores the factual foundation requirement

Some of the other problems with this approach, such as the actual deficiencies with trials and unexpected efficacies of plea bargains, should be explored more in 4.1.3: The apocryphal truth-seeking function of the trial

But that exact same defendant would not be any more or less guilty if that trial hadn't been held

Nor is that same apparent contradiction present when considering a defendant who pleads not guilty, hoping to succeed on an evidentiary issue, knowing full well that they are in fact guilty of the offence they're charged with

A witness who testifies in court stating that they did not do something that they in fact did is guilty of perjury, and may be charged with such

That same witness, as a defendant, may plead not guilty to something that they did in fact do without being guilty of perjury, or chargeable on that offence

Similarly, a defendant who pleaded guilty to a charge, only to be cleared with new evidence, would not have to worry about being prosecuted for their wrongful conviction

A judge may very well eviscerate a witness who said they didn't do something they did, but would be entirely out of line to do so for a defendant who pleaded not guilty to a crime they in fact committed

One would expect that if a plea were to be taken as evidence, it would need to be made under oath

Similarly, because a not guilty plea is understood as being something other than a declaration with truth content, counsel who allow or even encourage their clients to plead not guilty are not thought of as misleading the court

The way that the courts treat not guilty pleas is a clear signal that pleas are not simple declarations, and shouldn't be evaluated as such

There isn't anything inherently contradictory about a person entering a plea

\subsubsection{Conclusion}

It's necessary to distinguish between the different roles that different types of statements have to play in legal proceedings, as well as the "truths" that they're intended to convey

Just as a not guilty plea is not a declaration that the defendant is not, in fact, guilty, a guilty plea should not be understood as a declaration that the defendant is, in fact, guilty

When a defendant pleads not guilty, a trial is held to determine the question

When a defendant pleads guilty, a factual foundation is presented

A guilty plea is what the statute says it is: an admission of the essential elements of the offence and an invitation for the court to convict

Likewise, a not guilty plea signals to the court that none of the essential elements of the offence are admitted, and invites the prosecution to prove its allegations beyond a reasonable doubt


\subsubsection{Separating legal facts from legal fictions}

In most respects, the law is a relatively straightforward enterprise

But occasionally, legal problems arise that require abstract solutions that, at first blush, appear obtuse

Legal fictions are a means our law has used, and continues to use, in order to effect these solutions

These fictions allow the various actors in the justice system to make decisions and reach conclusions with the benefit of some assumptions in common

Some examples in criminal law

Reasonable person test

Foundation of many tests within the criminal law, especially as it concerns police and their grounds to detain, search, and arrest suspects without a warrant (or apply to a justice to obtain one)

Presumptions

Presumption of innocence

Deference to the trial judge

Presumptions about judges and juries

Public perception of justice

Key to decisions made about excluding evidence or releasing defendant's on bail

Not guilty plea

As discussed at some length in 4.1.1: The apparent untruthfulness of no-contest pleas

Equal treatment for equally-situated defendants

The practical impossibility of this is discussed further in 4.2.3: The trial penalty

In @bowersPunishingInnocent2007a, Josh Bowers recommends recasting equivocal no-contest pleas as legal fictions that have the actual effect of quickly curtailing wrongful pre-trial incarcerations

The potential effect of allowing factually innocent defendants to plead no-contest to end wrongful incarcerations without being entered on their criminal record (eg the rule in Rhode Island)

Asking the court to proceed as if the accused had committed the offences they're charged with, in the manner alleged by the prosecution, is just another of these fictions

\subsection{The presumption of innocence and the prosecutorial burden of proof}

The presumption of innocence and the prosecutor's burden of proving an offence beyond a reasonable doubt are hallmarks of the common law justice system. Legal professionals and academics see these concepts as indispensable tools for fighting wrongful convictions. When properly implemented, these concepts place a significant burden on the prosecutor, so only those who are guilty of an offence stand any chance of being convicted. The joint operation of these concepts will inevitably result in some guilty persons walking away without conviction. However, this sacrifice is generally considered worth the added protections afforded to the truly innocent. 

These legal concepts seem to run contrary to the practice of non-inculpatory self-convictions. Where the prosecutor has the burden of proving the charges and the court must consider the defendant innocent until the prosecutor discharges that burden, it appears contradictory and unjust to convict a defendant who maintains their innocence either tacitly or explicitly.\footnote{Cite to one or more articles that take this position. Alschuler gets into this briefly, but the author who discusses Alford's ``sworn testimony" at length is the source here.} In cases where the defendant maintains their innocence, it superficially follows that they continue to be presumed innocent until the prosecutor discharges their burden. 

Although this argument appears to have some legs, it is ultimately baseless, as it overlooks the following key factors:

\begin{itemize}
    \item \textbf{Admissions.} Defendants are entitled to make admissions against their legal interests, which may obviate the presumption of innocence.
    \item \textbf{Factual foundation.} Where non-inculpatory self-convictions are permitted, courts typically must be satisfied that there is a factual foundation for the conviction apart from the plea itself. This stipulation generally mirrors the same requirement for inculpatory self-convictions.\footnote{Although this is true for most jurisdictions, it is not universally so. For example, \hl{X-state} requires a factual foundation for guilty pleas, but not for \textit{nolo contendere} pleas.} 
    \item \textbf{Judicial discretion.} Although Canadian courts must accept not guilty pleas on their face, judges have ultimate discretion over whether to allow a defendant to self-conviction.\footnote{Again, although this is true for most jurisdictions, it is not universally so. For example, \hl{Y-state} insists that courts must accept \textit{nolo contendere} pleas. Similarly, the \textit{nolo contendere} ``plea procedure" used in Canadian courts is arguably beyond judicial discretion, as discussed in \hl{Z-section}.}
\end{itemize}

\subsubsection{Admissions}

\textit{Criminal Code} s 655 allows defendants to admit any facts agreed to by the prosecutor. Typically, these facts will support or consist entirely of an element of the offence the defendant is with, though any fact that both parties agree on can be admitted under this section. Just as a guilty plea obviates the need for the prosecutor to prove the offence, these formal admissions remove any need for either side to prove the fact.\footnote{See @2009mbca37 for a discussion of the distinction between formal and informal admissions.} The legislation does not require a judicial inquiry into the factual foundation.\footnote{Some jurisdictions have established a common law basis for judges to inquire into the facts admitted and accept or reject them based on the outcome of that inquiry. See @2019bcsc2048 at para 7.} Absent this oversight and any directive to the contrary, \textit{Criminal Code} s 655 does not require a defendant to have committed the essential elements of the offence or to admit that they did commit the essential elements of the offence. Facts that are admitted and later found to be contradicted by evidence may nevertheless remain proven.\footnote{example?}

Prosecutors and defendants may only use \textit{Criminal Code} s 655 where they both agree that the fact in question is admissible.\footnote{See the old SCC case re formal admissions where the defendant tried to sneak one in without the Crown's approval. See ยง3.x above.} That prosecutors and defendants use this section strongly suggests that it benefits both.

Defendants may wish to forego proof of some or all of the facts of their case for several reasons. In some cases, there is no reason to doubt that the prosecution will prove certain offence elements. A prosecutor is likely to be able to prove elements like date, time, jurisdiction, and identification in a domestic violence prosecution without difficulty. Agreeing to these facts saves court time and judicial attention, which can be minimal resources in busy court circuits. Admissions can also be advantageous in situations where the evidence is less certain to be proven. For example, when the prosecution wishes to call a witness whose evidence will inevitably harm the defendant to one degree or another, the defendant may wish to have the witness excused in exchange for a manageable set of agreed facts. In cases where the only genuine dispute lies with a legal issue, both parties may agree to have all evidence admitted by consent and limit trial time to arguing that issue. 

Admissions are a way to signal to the court what the parties do and do not need to demonstrate. Each admission dispenses with the presumption of innocence and burden of proof for the fact admitted. Where the parties admit facts but decide to call evidence, witnesses, including the defendant themselves, may testify in a way that is inconsistent with those facts. However, the facts are nonetheless admitted, thus dispensing the need for proof. No-contest pleas admit \textit{all} of the essential elements needed to prove the offence and dispense with the presumption of innocence across all aspects of the offence charged.

The problem with this argument generally is that the presumption of innocence is the defendant's to waive. Like any other person, criminal defendants are susceptible to ``buyer's remorse," minimization, outright denial, and misunderstanding of legal concepts like ``self-defence." There is nothing seemingly incongruous or unjust about a defendant who waives this presumption, only to resile in one form or another later in the proceedings.

\subsubsection{Factual foundation}

Notwithstanding that defendants can legally admit facts against their interests, concerns that they might do so inappropriately may reasonably remain. While there is nothing implicitly wrong with a defendant admitting a fact against their interests, there may be problems with defendants admitting baseless facts against their interests. There may also be problems with defendants pleading guilty to charges whose underlying facts do not meet the legal threshold for a conviction. For these reasons, common law courts must usually verify the offence's factual foundation. In Canada, judges must be satisfied that the facts support the charges.

The facts underlying both the \textit{Alford} and \textit{Hector} decisions are instructive in this respect. In \textit{Alford}, the defendant 

\textit{Alford} and \textit{Hector} both show that 

Nothing prevents legislatures from requiring judges to more rigorously examine the factual foundation of non-inculpatory self-convictions than they would be required to do for inculpatory pleas. 

\subsubsection{Discretion}

Judges still have discretion over whether to reject or accept self-convictions.

\subsection{Fairness}