\section{Guilty: the inculpatory plea}

Since the first \textit{Criminal Code of Canada},\footnote{\textit{The Criminal Code}, 1892 (55-56 Vict), c 29 [\textit{Criminal Code, 1892}].} criminal defendants have been able to self-convict rather than take their charges to trial.\footnote{See \textit{ibid} s 657.} This early provision did not limit a defendant's ability to enter a guilty plea and did not give judges discretion to accept or reject it. Over time, Canadian common law delineated a defendant's ability to plead guilty and stipulated that judges may only accept knowing, voluntary, and unequivocal guilty pleas. Canadian Parliament eventually codified a form of this common law requirement as \textit{Criminal Code} s 606(1.1) in 2002.\footnote{See the \textit{Criminal Law Amendment Act}, 2001, SC 2002, c 13, s 49.} In Canada, the guilty plea is the most frequently used uncontested plea.\footnote{See note XXX above (the note referencing ONCJ's stats).}

\subsection{The guilty plea historically}

The modern guilty plea evolved from a longstanding practice where courts allowed defendants to confess their crimes without requiring the state to prove them. Before the 11\textsuperscript{th} century Norman conquest, the Anglo-Saxons of England established a procedure for confessions to criminal charges.\footnote{See Albert W. Alschuler, ``Plea Bargaining and Its History" (1979) 79:1 Colum L Rev 1 at 7.} Confessions were rudimentary and risky, lacking the modern guilty plea's procedural protections and finality.\footnote{See \textit{R v Baird}, 1908 CarswellSask 23 at para 2 [\textit{Baird}].} There is some question over how favourably the judiciary looked on these pleas and how often they were prepared to accept them.\footnote{See Alschuler, \textit{supra} note XXX at 5 - 6. Although Alschuler asserts that guilty pleas were ``discouraged" and met with ``general disapproval", none of my research into guilty pleas entered in Canada revealed any similar concerns amongst Canadian judges. For his part, Alschuler does not cite any cases, historical studies, or sources of any kind to support this claim. This by no means excludes the possibility that Canadian criminal law actively discouraged guilty pleas once upon a time, but if it did, this attitude was not reflected in the written decisions discussing guilty pleas.} However, toward the end of the 19\textsuperscript{th} century, courts began regularly reporting cases involving guilty pleas.\footnote{See \textit{ibid}, \textit{supra} note XXX at 7 - 8.} By the 1920s, criminal justice surveys in America revealed that many more defendants preferred to plead guilty, and the guilty plea became fixed in American and Canadian criminal proceedings.\footnote{See \textit{ibid}, \textit{supra} note XXX at 26. The percentages of guilty pleas Alschuler reports range between 76 - 90\% in several major American cities at the time.}

\subsection{Plea voluntariness and comprehension}

As defendants entered guilty pleas more commonly, procedural protections around them increased. One of the most significant was the \textit{plea voluntariness and comprehension inquiry}. To the extent that defendants were presumed innocent and entitled to remain silent, courts examined defendants to ensure that they only waived those procedural protections knowingly and voluntarily. This safeguard, referred to here and throughout as the \textit{plea inquiry}, exists to ensure that defendants know their rights when charged with an offence and that they only waive those rights of their own free will. Not all jurisdictions require judges to review the same sets of factors or place the same importance on a judge's review of the same with defendants. But despite these emphatic differences, the broad factors that courts consider are generally similar, and may include:

\begin{itemize}
\item \textbf{Voluntariness and knowledge.} Defendants entering pleas must do so voluntarily and know the nature and consequences of their plea. As early as the 16\textsuperscript{th} century, English courts required that defendants enter guilty pleas voluntarily,\footnote{See Alschuler, \textit{supra} note XXX at 12 - 15. Alschuler attributes this practice to the co-evolution of the plea alongside the common law confessions rule. As with confessions, guilty pleas allow defendants to self-convict. However, unlike confessions, guilty pleas are presumed valid and admissible as evidence.} Whether defendants enter a plea voluntarily and knowingly impacts whether they were treated \textit{fairly} in being asked to enter it.
\item \textbf{Factual basis.} In addition to ensuring that defendants enter guilty pleas knowingly and voluntarily, local laws may also require judges to verify that there is some factual basis for the plea such that there is enough evidence to sustain a conviction.\footnote{Contrast this requirement with the common-law rule that judges need not inquire into the factual circumstances of \textit{nolo contendere} pleas. See George Edward Pickle, "Nolo Contendere Pleas in Criminal Tax Cases" (1975) 13:2 Am Crim L Rev 249 at 250 - 251.} A sentencing judge's obligation to confirm a factual foundation for the plea ensures that the allegations are reviewed.\footnote{Judges are not expected or allowed to operate as independent fact-finders. See \textit{R v Woroniuk}, 2019 MBCA 77, where the Manitoba Court of Appeal reversed a ``flagrant error in law" committed by sentencing judge Brian Corrin PCJ who took it upon himself to investigate the factors counsel laid out at sentencing. As a result, the scope of the judge's inquiry into the charge's factual foundation is limited, but the inquiry is nevertheless required.} The factual foundation requirement helps guarantee that uncontested pleas are \textit{truthful} and accurate.
\item \textbf{Collateral consequences.} ``Collateral consequences" are consequences that the court does not intentionally impose but nonetheless directly result from criminal convictions. These consequences can be legal or extra-legal. Legal collateral consequences result from intersections between different legislative schemes, where a decision made under one legislative scheme has consequences under another.\footnote{These consequences may include extended driving restrictions imposed by the provincial licensor (rather than the court), restrictions on purchasing certain medications or other substances, and cancelled foreign work permits and permanent resident designations, among others. Many American jurisdictions have codified requirements that courts alert defendants of the inevitable collateral consequences of pleading guilty to an offence. Requirements that courts tell defendants about collateral gun ownership and immigration consequences are standard. For American jurisdictions, the National Inventory of Collateral Consequences of Conviction maintains a comprehensive online database of these consequences. See the National Inventory of Collateral Consequences of Conviction, online: \textless https://niccc.nationalreentryresourcecenter.org/\textgreater} Extra-legal collateral consequences refer to the non-legal consequences that flow from the defendant's offending behaviour.\footnote{These may include breakdowns in social or familial relationships, opportunity losses, or physical injuries, among many others. See \textit{R v Suter}, 2018 SCC 34, [2018] 2 SCR 496 [\textit{Suter}]. I discuss this decision in detail immediately below.} Collateral consequences go to a plea's \textit{fairness}. When defendants enter a plea knowing about the collateral consequences of doing so, their pleas are knowledgable and voluntary. But when defendants enter pleas without knowing the consequences of doing so, the opposite is true.
\end{itemize}

\subsection{Guilty pleas in Canada}

Before Parliament codified Canadian criminal law in 1892, reported decisions dealing with guilty pleas were rare.\footnote{My research on the guilty plea's history in Canada only revealed two such decisions: namely, \textit{R v Morrison}, 1879 CarswellNB 35 and \textit{R v Morin}, 1890 CarswellQue 17.} However, following codification, reported cases dealing with guilty pleas abounded and Canadian guilty plea procedures adapted in turn. In the early days of Canadian criminal law, much remained unsettled about guilty pleas, their nature, and what legal consequences stemmed from entering one. As cases made their way through the justice system, courts settled on predictable rules and uniform procedures that made it possible for defendants to know the consequences of waiving their right to trial much better.\footnote{Early on, questions such as whether a guilty plea waived a right to trial were still issues when Parliament codified Canadian criminal law. See \textit{R v Davidson}, 1992 CanLII 2570; see also \textit{R v Gillis}, 1914 CarswellYukon 6, where the court treats the subject as a viable but generally settled question of law.}

\subsubsection{Voluntariness and knowledge}

Before codification, Canadian courts appeared to operate on the principle that a person pleading guilty must be presumed to know the law.\footnote{See \textit{Morrison} at paras 9, 13; \textit{Morin} at para 145. The courts in both cases cite this principle. However, given that these were the only two Canadian pre-codification decisions I could find that covered guilty pleas, there is little to draw from their similarity in this respect.} However, developments in the following decades moved towards ensuring that defendants better knew the implications of self-convicting before entering uncontested pleas. As the 20\textsuperscript{th} century unfolded, Canadian courts began insisting that defendants only enter guilty pleas knowingly and voluntarily and refusing guilty pleas from defendants induced or threatened to self-convict.

Plea voluntariness ensures that defendants who enter guilty pleas do so of their own free will in an oppression-free environment and that a failure to ensure these conditions obtain may warrant appellate intervention. For example, in \textit{R v Baird},\footnote{See \textit{R v Baird}, 1908 CarswellSask 23.} the Saskatchewan District Court heard an appeal from a landowner convicted of allowing a fire to spread from his land to a neighbour's property. Baird, the landowner, pleaded guilty but argued on appeal that he had entered his guilty plea under oppression. The court reviewed the proceedings below but found no evidence that the justice who took Baird's plea had acted oppressively or had induced him to plead guilty. The court dismissed his appeal. 

Canadian courts similarly required defendants entering guilty pleas to know the legal consequences of doing so. In \textit{R v Tom},\footnote{\textit{R v Tom}, 1928 CarswellNS 10.} the defendant pleaded guilty to possessing narcotics and was sentenced to six months of jail and a \$500 fine. On appeal, Tom argued that he had not understood the consequences of pleading guilty. His predicament was compounded by the fact that the prosecutor promised Tom he would only receive a small fine if he did so. A majority of the Nova Scotia Supreme Court allowed the appeal. The dissent agreed that there seemed to be a miscarriage of justice but disagreed that the court had jurisdiction to order a remedy. 

\subsubsection{Factual basis}

Canadian criminal courts have also long required a \textit{factual foundation} as a prerequisite for accepting a guilty plea. This requirement reflects the common law's perennial concern for preventing wrongful convictions\footnote{See \S 4.1.2 below for my more detailed overview and discussion of wrongful convictions.} by ensuring that the crimes alleged can actually be made out on the allegations, and that there are no allegations that the defendant wishes to contest. At the turn of the 20\textsuperscript{th} century, the Ontario Supreme Court had the opportunity to review an unusual set of facts that exemplified this principle. In \textit{R v Herbert},\footnote{See \textit{R v Herbert}, 1903 CarswellOnt 829} the Ontario Supreme Court reviewed a case where the defendant, Herbert, pleaded guilty to a murder nobody suspected he had committed. After pleading guilty, Herbert implicated a co-defendant, Sifton, who pleaded not guilty and was acquitted after a trial. 

During this time, Herbert developed misgivings about his guilty plea and looked to withdraw it. At the time in Canada, the death penalty was the mandatory sentence for Herbert's crime. The Ontario Supreme Court did not review the allegations brought out at Sifton's trial, but nonetheless concluded that Herbert's guilt was ``absolutely inconsistent" with Sifton's acquittal, and that there was therefore ``no theory that can be suggested upon which [the co-defendant] could be innocent and [Herbert] guilty."\footnote{See \textit{ibid} at paras 4 - 6.} That is, the allegations did not logically support the charges. Absent any possible factual foundation for the plea, the court allowed Herbert to withdraw it.

\subsubsection{Collateral consequences}

Canadian courts recognize \textit{legal and extra-legal collateral consequences} as valid sentencing considerations. \textit{Legal collateral consequences} arise when a legal outcome obtained under one statute or legislative scheme triggers a provision in another. Increased insurance premiums after an impaired driving conviction, losing the ability to purchase firearms after a violent offence conviction, or revoking a work visa after a theft from an employer conviction are all legal collateral consequences following criminal convictions. Although Canadian courts may consider these consequences when sentencing a defendant, they are not explicitly required to do so. 

Collateral immigration consequences are an especially well-litigated example of this problem, with \textit{Wong} providing a recent and authoritative example. In \textit{Wong}, the defendant pleaded guilty to a cocaine trafficking charge but did not know that doing so would affect his Canadian residency. He appealed to withdraw his guilty plea. The court agreed that defendants must know the ``legally relevant collateral consequences" of pleading guilty for the plea to be valid, but split on which legal test they should apply.\footnote{The majority held that a defendant must be able to show subjective prejudice to withdraw an uninformed guilty plea of this variety. The dissent suggested accomplishing this through a modified objective test.} Although both agreed that Wong entered his guilty plea without sufficient information, the majority dismissed his appeal while the dissent would have allowed it.

Canadian courts also recognize extra-legal collateral consequences as a valid sentencing consideration, as seen in the \textit{Suter} decision.\footnote{See \textit{Suter}, \textit{supra} note XXX at paras 45 - 59.} While arguing in a parking lot with his wife after dinner and a drink, Suter drove onto a restaurant patio and accidentally killed a two-year-old child. Although there was no other evidence that Suter was impaired, the police demanded a breath sample, given the circumstances surrounding the collision. Having spoken with a lawyer who unwisely advised him not to provide a sample, Suter refused to do so when required. As a result, the police charged him with refusing to provide a breath sample in a case involving death. While his case was still pending, a group of vigilantes who were angry about the offence attacked him and cut off one of his thumbs. Suter eventually pleaded guilty and asked the judge to impose a non-custodial sentence.

At sentencing, the judge relied on the fact that Suter was the victim of a violent attack in when sentencing him, and held that leniency was in order. However, the judge did not feel that a non-custodial sentence was appropriate and sentenced Suter to four months of jail. Both he and the Crown appealed. The Alberta Court of Appeal ruled that the sentencing judge erred for several reasons, including relying on vigilante violence that did not ``emanate from state misconduct", and re-sentenced Suter to 26 months of custody. Suter appealed to the Supreme Court of Canada, who found that the Alberta Court of Appeal incorrectly dismissed Suter's attack and injuries as irrelevant. The Supreme Court of Canada ruled that collateral consequences may include ``any consequence arising from the commission of an offence, the conviction for an offence, or the sentence imposed for an offence, that impacts the offender." In Suter's case, this included the violent retributive attack he suffered for his role in the offence. The majority allowed the appeal and varied the sentence to time served. The lone dissenting judge would have upheld Suter's original four-month custodial sentence.

\subsubsection{Unequivocal}

Unlike other common law jurisdictions, Canadian common law requires that defendants who enter guilty pleas do so \textit{unequivocally}. An unequivocal plea is one that the defendant intended to enter. In cases where a reviewing or first-instance court is left wondering whether a defendant intended to plead guilty, the plea is equivocal and should not be accepted or upheld. This requirement appears to trace back to a comment from the dissent in \textit{Adgey v R}.\footnote{See \textit{Adgey v R},  1973 CanLII 37 (SCC), [1975] 2 SCR 426 at 433f [\textit{Adgey}].} In that case, the defendant Adgey pleaded guilty to several offences with duty counsel's assistance but wanted to explain some of the charges. Some of Adgey's explanations were consistent with innocence. The judge nonetheless accepted his guilty pleas without conducting a plea inquiry. Adgey appealed on the ground that the judge should not have done so, given that his explanation for the offences was consistent with innocence. 

The majority dismissed the appeal, finding that the judge was not bound to conduct a plea inquiry in the case and that Adgey had formally pleaded guilty. The dissent disagreed and argued that the pleas at first instance were equivocal as it was difficult to tell whether Adgey wanted to self-convict on all the charges he did. The requirement that pleas be unequivocal became entrenched in Canadian criminal law follows the formula expressed in \textit{Adgey}, despite being delivered by the dissent. Forty-five years later, the majority in \textit{Wong} endorsed this requirement, solidifying its place in Canadian criminal procedure

Canadian courts have interpreted this requirement in two main ways. The first applies the unequivocal requirement to the \textit{plea's content}. Under this approach, the defendant pleads unequivocally when they admit they were responsible for the offence.\footnote{See e.g. \textit{R v Smoke}, 2017 SKQB 345 at para 19: ``As to requirement that the plea is unequivocal, s. 606(1.1)(b)(i) says that the accused must understand and admit the essential elements of the offence"; } The second applies the unequivocal requirement to the \textit{defendant's subjective intent} to bring about the consequences of their plea. Under this approach, the defendant pleads unequivocally when they unreservedly acknowledge their intention to self-convict.\footnote{See e.g. \textit{R v Singh}, 2014 BCCA 373 at para 43: ''A plea is presumed to be unequivocal, meaning it is one that is in no way 'qualified, modified or uncertain'", citing \textit{R v T(R)}, 1992 CanLII 2834 (ON CA) at 521. See also \textit{R v Wiebe}, 2012 BCCA 519 at para 28.} Although the former position is broadly supported, the latter is more cogent. It follows closely with how \textit{Adgey} originally used the term, and it avoids any definitional overlap with the factual foundation requirement. More importantly, it reflects that a guilty plea's proof-content is its primary meaning and that this meaning must be unequivocal. As will be seen throughout this thesis, keeping a plea's proof-content separate from its truth- and belief-content is key to classifying and understanding them.